Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2021-01-27 doc/SA-Hermaphrodites-wDemography.tex
Update Results Section and Figures
Big commit: finished up first draft of results section, minor edits
to introduction. Also added new data for fig. 3, and a new fig. 4
with Mimulus simulations.

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{nameref}

@@

 Whether a population persists or goes extinct depends in large part on its ability to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, environmental shifts \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, or genetic constraints \citep{ConnallonHall2018}.

@@

%Conflicting selection and gene-flow between different classes of individuals, or genetic trade-offs between fitness components, can impose constraints on adaptation with particularly interesting evolutionary consequences \citep{CharlesworthHughes2000, ConnallonHall2018}. 
Genetic constraints can arise as a consequence of conflicting selection and gene-flow between different classes of individuals. That is, genetic trade-offs between fitness components can constrain adaptation \citep{CharlesworthHughes2000, ConnallonHall2018}. On the one hand, such constraints prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase the population's overall extinction risk \citep{holman2013consequences,harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences the population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the contribution of genetic constraints to the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Classical population genetic models consider relative fitnesses of genotypes, and therefore do not consider the population dynamical consequences of evolutionary change. To study the population dynamical consequences of evolutionary change, a model is required that links changes in genotype distributions to population dynamics.  The demographic processes of birth and death drive both changes in genotype distributions and population dynamics, and therefore provide a fundamental link between evolution and ecology \citep{metcalf2007evolutionary}.

In this paper, we use such an evolutionary demographic approach (model?) to study whether the parameter space predicted to be polymorphic by population genetics models is, indeed, demographically viable. We apply the model to study the maintenance of polymorphisms through sexually antagonistic selection in both dioecious and hermaproditic species. A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

%Such an eco-evolutionary model requires a map from genotype to phenotype, from phenotype to individual demographic rates (birth and death rates), and finally from demographic rates to population level characteristics \citep{coulson2006putting}.    
%Genetic trade-offs between fitness components due to pleiotropy represent a form of genetic constraint with particularly interesting evolutionary consequences because they act as a double-edged sword
%The nature and extent of genetic variation observed in natural populations should therefore reflect the balance between the contribution of those constraints to the maintainance of genetic polymorphism and the demographic consequences of the resulting maladaptation.

Sexually antagonstic selection, due to genetic trade-offs between male and female fitness, are a common feature of sexually reproducing populations, and are thought to contribute not only to the maintenance of genetic variation, but also the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, "sexually antagonistic" selection (abbreviated SA hereafter) can arise whereby beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Such alleles with opposing fitness effects through male and female sex functions create analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

@@

Following convention, we parameterize the relative fitnesses of each genotype through each sex function ($w_{i}$ and $w^{\prime}_{i}$) to be bounded by $(0,1)$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}). The SA locus does not affect survival and transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\sigma$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_A
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

@@

\begin{table}[htbp]
 \centering
 \caption{\bf Fitness expressions for Sexually Antagonistic selection in hermaphrodites ()}
\label{tab:Fitness}
\begin{tabular}{lccc}
 \toprule
					&  \multicolumn{3}{c}{{\textit{Genotype}}} \\ 
\cline{2-4}
					& $AA$			& $Aa$ 					& $aa$ 		\\ \hline
Female function ($w_{i}$):	& $1$		& $1 - h_f s_f$	& $1 - s_f$ \\	
Male function ($w^{\prime}_{i}$):		& $1 - s_m$& $1 - h_m s_m$	& $1$ 		\\	
\hline
\end{tabular}
\end{table}

@@

Unless stated otherwise, we use the following parameter values for the demographic rates in the model: $\sigma_j = \sigma_A = 0.6$ and $\gamma = 0.05$, which are similar to those used in \citet{deVriesCaswell2019b}. Our parameters of interest include fertility, $f$, the inbreeding depression parameters, $\delta$, $\delta_j$, $\delta_a$, $\delta_{\gamma}$, and the selection parameters $h_f$, $s_f$, $h_m$, and $s_m$, which are given different values for each analyses as described in the figure captions.

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Analyses} \label{sec:analyses}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Demographic and fitness data for {\itshape M. guttatus}} \label{sec:MguttMethods}

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized our model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic matrix data made publicly available on the demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix B.x}.

We used extensive demographic rate data reported for the 'Eagle Meadows' population in Stanislaus National Forest (California, USA) studied by \citet{PetersonEtAl2016}. This study entailed a large common garden experiment which took place at the 'Eagle Meadows' site. We used demographic rates estimated for this population grown in its native habitat (i.e., we did not use data for non-local populations grown at Eagle Meadows). \citet{PetersonEtAl2016} collected data for numerous life-history stages and transition rates; for the Eagle Meadows population these estimates were as follows: seed bank survival ($D = 0.534$), seed germination rate ($G = 0.469$), flower production ($F = 0.64$), ovules per flower ($O = 614)$, seedling recruits proportional to clonal rosette recruits ($A = 6.7e-4$), overwinter survival ($S = 0.179$), rosette production ($R = 8.71$) (see corrected Tables.~1 and S2 in \citealt{PetersonEtAl2017}). These data were also available through the Compadre demographic matrix database \citep{CompadreDB2020}. The resulting transition matrices for this population involved three life-history stages ($\omega = 3$; seed, seedling, and rosette), and individual elements of the transition matrix ($\tilde{\mbf{A}}$) were calculated as products of the above rates (see Matrix 1 in \citealt{PetersonEtAl2016}).

Unfortunately, estimates of selfing rates and inbreeding depression are not also available for the Eagle Meadows population of {\itshape M.~guttatus}. However, selfing rate estimates are available for a variety of other populations, and vary in magnitude from near complete outcrossing to predominant selfing ($C \approx 0$ to $0.75$; \citealt{RitlandGanders1987, Ritland1990, Willis1999b}). Additionally, estimates of inbreeding depression at several of the life-history stages/fitness components that were included in the data of \citet{PetersonEtAl2016} are available for two intensively studied populations in the Cascade Mountains of Oregon (Iron Mountain and Cone Peak) \citet{Willis1993, Willis1999a, Willis1999b}. Using the data provided in \citet{Willis1993}, we estimated the proportional decrease due to inbreeding depression in seed germination rate ($\delta_{G} = 0.085$), flower number ($\delta_{F} = 0.2$), and overwinter survival ($\delta_{S} = 0.38$).

Using these combined demographic rates, selfing rate, and inbreeding depression estimates, we constructed a corresponding stage $\times$ genotype mendelian matrix model with a single SA locus affecting female and male fertility (as described above). Using this empirically parameterized model, we are able to make predictions about the scope for demographically viable SA polymorphism at a single locus in a hypothetical population with the same demographic rates as observed in the Eagle Meadows population, for a range of selfing and inbreeding depression rates observed in other natural populations. However, this should not be confused with an explicit prediction for the likelihood of SA polymorphism in the Eagle Meadows population.

Interestingly, a polymorphic chromosomal inversion (Inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. Inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production which varies among years. The genetic basis of the fitness effects caused by Inv6 are not yet known, but the net result is a "supergene" with uniquely strong fitness effects that segregates as a single diallelic locus. We estimated dominance and selection coefficients for the effect of Inv6 on pollen viability and flower production from greenhouse crossing experiments using mapping populations and field estimates. Across four mapping populations, the effect of Inv6 on pollen viability was partially recessive ($h_m \approx 0.35$). Given this average dominance, the selection coefficient in the field is approximately $s_m = 0.88$. It was not possible to estimate the average dominance of the effect of Inv6 on flower production, so we conservatively assume that it was additive ($h_f = 0.5$). Based on this assumption, the selection coefficient Inv6 on flower production in the field was $s_f = 0.384$ in 2012, and $s_f = 0.232$ in 2013. As a final proof-of-concept test of our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, Inv6 appears to fall in demographically viable SA polymorphic parameter space, as might be expected given its observed frequency in the Iron Mountain population.

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%%%%%%%%%%%%%%%%%%%%%%%%
\ldots

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polymorphism and Extinction} \label{sec:Res:PolyExt}

 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{../output/figs/extinctionThresholdsFig.pdf}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the mendelian stage-structured matrix model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). We are particularly interested in identifying and quantifying "demographically viable polymorphic parameter space", which corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity with $h = 1/2$, and dominance reversal with $h = 1/4$); extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel).}}
 \label{fig:extThresholds}
 \end{figure}

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polymorphism in the absence of inbreeding depression} \label{sec:Res:PolyNoID}

 \begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/polymorphicSpaceNew.pdf}
 \caption{\footnotesize{Proportion of demographically viable parameter space (out of total $s_f \times s_m$ space with $\max(s) = 0.15$) in the absence of inbreeding depression (i.e., assuming $\delta = \delta_i = 0$, where $i \in \{j,a,\gamma\}$), plotted as a function of the population selfing rate. Results are shown for three fertility values corresponding to low, medium, and high fertility (blue, green, and red points respectively) under additive (panel A; $h = 1/2$), and partially recessive (panel B; $h = 1/4$) SA fitness effects. Each point was calculated by numerical integration of the corresponding SA invasion conditions and extinction threshold predicted by the mendelian matrix model (see \hyperref[sec:analyses]{Analyses} section), while solid lines were produced by numerically integrating the analytic expressions for the single-locus invasion conditions from the population genetic models of \citet{JordanConnallon2014} and \citet{Olito2017} (solid black lines).}} 
 \label{fig:polySpace}
 \end{figure}

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Demographic effects of inbreeding depression} \label{sec:Res:InbreedEffects}

 \begin{figure}[htbp]
 \centering
 \includegraphics[scale=0.75]{../output/figs/deltaPolymorphicSpaceTitrate.pdf}
 \caption{\footnotesize{Effects of early- and late-acting inbreeding depression on the proportion of demographically viable parameter space (out of total $s_f \times s_m$ space with $\max(s) = 0.15$), plotted as a function of the population selfing rate. In all plots, the strength of inbreeding depression decreases as the selfing rate goes up following a simple model of purging recessive deleterious mutations (see \hl{Appendix X.x} for details). Only single, $\delta_i$ terms indicated in the legend ($i \in \{j,a,\gamma\}$) are allowed to vary at once (all others are set to $0$). Effects of early-acting inbreeding depression are shown on the left (panels A, C; $\delta$) and for late-acting inbreeding depression on the right (panels B, D; $\delta_i$) for additive ($h = 1/2$) and partially recessive ($h = 1/4$) SA fitness effects, respectively. All results used the same 'high female fertility' value of $f = 6.5$ (see Fig.\ref{fig:polySpace}). Each point was calculated by numerically integrating the corresponding SA invasion conditions and extinction threshold predicted by the model (see \hyperref[sec:analyses]{Analyses} section), while solid lines were produced by numerically integrating the analytic expressions for the single-locus invasion conditions from the population genetic models of \citet{JordanConnallon2014} and \citet{Olito2017} (solid black lines). The population genetic model predictions for varying $delta$ are plotted on both left- and right-hand panels for comparison, and to highlight the different y-axes scales.}} 
 \label{fig:deltaPolySpace}
 \end{figure}

>>>>

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
\usepackage[colorlinks=true, allcolors=black]{hyperref}

@@

 Whether populations persist or go extinct depends heavily on their ability to adapt to the environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, environmental shifts \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, or genetic constraints \citep{ConnallonHall2018}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

@@

%Conflicting selection and gene-flow between different classes of individuals, or genetic trade-offs between fitness components, can impose constraints on adaptation with particularly interesting evolutionary consequences \citep{CharlesworthHughes2000, ConnallonHall2018}. 
Fitness trade-offs constraining adaptation have particularly interesting evolutionary consequences. On the one hand, they prevent individuals (or classes of individuals) from reaching their phenotypic optimum in one or more fitness components, which can increase the population's overall extinction risk \citep{holman2013consequences, harts2014demography}. On the other hand, they provide an effective mechanism for the maintenance of genetic variation, which influences the population's capactiy for future adaptation \citep{Fisher1930, CharlesworthHughes2000, ConnallonHall2018, MatthewsConnallon2019}. Hence, for traits under conflicting selection, the nature and extent of genetic variation observed in natural populations should reflect a balance between the contribution of genetic constraints to the maintainance of genetic polymorphisms, and the population dynamical consequences of the resulting maladaptation.

Classical population genetic models consider relative fitnesses of genotypes, and therefore do not consider the population dynamical consequences of evolutionary change. To take the population dynamical consequences into account, a model is required that links changes in genotype distributions to population dynamics.  The demographic processes of birth and death drive both changes in genotype distributions and population dynamics, and therefore provide a fundamental link between evolution and ecology \citep{metcalf2007evolutionary}.

In this paper, we use an evolutionary demographic model to study when the parameter conditions predicted to be polymorphic by population genetics models are also demographically viable. We apply the model to study the maintenance of polymorphisms by sexually antagonistic selection in both dioecious and hermaproditic species. A major strength of the model is that we can use empirically estimated demographic rates to parameterize the model, enabling us to make predictions about the scope for demographically viable SA polymorphisms that are grounded in the biology of real populations. We demonstrate this with a case study of {\itshape Mimulus guttatus} (now {\itshape Erythranthe guttata})

%Such an eco-evolutionary model requires a map from genotype to phenotype, from phenotype to individual demographic rates (birth and death rates), and finally from demographic rates to population level characteristics \citep{coulson2006putting}.    
%Genetic trade-offs between fitness components due to pleiotropy represent a form of genetic constraint with particularly interesting evolutionary consequences because they act as a double-edged sword

Sexually antagonstic selection, due to genetic trade-offs between male and female fitness, are a common feature of sexually reproducing populations, and are thought to contribute not only to the maintenance of genetic variation, but also the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, "sexually antagonistic" selection (abbreviated SA hereafter) can arise whereby beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

@@

Following convention, we parameterize the relative fitnesses of each genotype through each sex function ($w_{i}$ and $w^{\prime}_{i}$) to be bounded by $(0,1)$, with dominance and selection coefficients $h_f, s_f$ and $h_m, s_m$ determining the decrease in fertility through each sex function relative to the most fit genotype ($AA$ has highest female fertility, $aa$ the highest male fertility; see Table \ref{tab:Fitness}). The SA locus does not affect survival and transition rates. However, the survival matrices can be used to model the fitness effects of inbreeding depression at later stages of development by allowing different stage-specific survival and transition rates for individuals produced by self-fertilization vs.~outcrossing. By contrast, the parameter $\sigma$ only affects inbreeding depression through viability of selfed ovules. With this in mind, we define survival matrices for individuals produced by selfing and outcrossing as follows:
\begin{linenomath*}
\begin{equation} \label{eq:US}
	\mbf{U}^S = \left(
					\begin{array}{cc}
						\sigma_j (1 - \delta_j) \big(1 - \gamma (1 - \delta_{\gamma}) \big) & 0 \\
						\sigma_j (1 - \delta_j) \gamma (1 - \delta_{\gamma})      & \sigma_a (1 - \delta_a)
					\end{array}
				\right) \\
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation}\label{eq:UX}
	\mbf{U}^X = \left(
					\begin{array}{cc}
						\sigma_j(1 - \gamma) & 0 \\
						\sigma_j \gamma      & \sigma_a
					\end{array}
				\right)
\end{equation}
\end{linenomath*}

@@

Throughout our analyses, we distinguish between early- and late-acting inbreeding depression. We quantify early-acting inbreeding depression using $\delta$, and late-acting inbreeding depression using $\delta_i$ (where $i \in \{j,a,\gamma\}$). An important difference between early- and late-acting inbreeding depression in the model is that $\delta$ does not affect $\mbf{U}^S$, the survival matrix for offspring produced by selfing (see Eq.~\ref{eq:Atilde}), while the $\delta_i$ terms directly alter the stage-specific survival and transition rates in $\mbf{U}^S$ (see Eq.~\ref{eq:US}).

@@

\begin{table}[htbp]
 \centering
 \caption{\bf Fitness expressions for Sexually Antagonistic selection ($w_{i}$)}
\label{tab:Fitness}
\begin{tabular}{lccc}
 \toprule
					&  \multicolumn{3}{c}{{\textit{Genotype}}} \\ 
\cline{2-4}
					& $AA$			& $Aa$ 					& $aa$ 		\\ \hline
Female function ($w_{i}$):	& $1$		& $1 - h_f s_f$	& $1 - s_f$ \\	
Male function ($w^{\prime}_{i}$):		& $1 - s_m$& $1 - h_m s_m$	& $1$ 		\\	
\hline
\end{tabular}
\end{table}

@@

Unless stated otherwise, we use the following parameter values for the demographic rates in the model: $\sigma_j = \sigma_a = 0.6$ and $\gamma = 0.05$, which are similar to those used in \citet{deVriesCaswell2019b}. Our parameters of interest include fertility, $f$, the inbreeding depression parameters, $\delta$, $\delta_j$, $\delta_a$, $\delta_{\gamma}$, and the selection parameters $h_f$, $s_f$, $h_m$, and $s_m$, which are given different values for each analyses as described in the figure captions.

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Analyses} \label{subsec:analyses}

@@

For analyses of the demographic effects of inbreeding depression, we made two main simplifying assumptions: First, we acknowledge that if inbreeding depression is caused primarily by recessive deleterious mutations (as suggested by empirical data), it should covary negatively with the population selfing rate due to purging (\citealt{Charlesworth2009}; though we note that other processes could give rise to this pattern, e.g., \citealt{CrnokrakBarrett2002, Charlesworth2009,HedrickGarcia-Dorado2016}). Following \citet{Olito2019}, we incorporate such negative covariance by constraining the inbreeding depression terms in the model ($\delta$ and $\delta_i$, where $i \in \{J,A,\gamma\}$) to follow a simple declining function of the selfing rate: $\delta = \delta^{\ast} (1 - b (1 - L))$, where $\delta^{\ast}$ is the hypothetical severity of inbreeding depression for a completely outcrossing population, $b$ is a shape parameter determining how far $\delta$ will decline under complete selfing (when $C = 1$), and $L$ is an expression describing the expected deleterious mutation load due to recessive deleterious mutations that includes an additional shape parameter, $a$, which determines the curvature of the overall function for $\delta$ (see Appendix E in \citealt{Olito2019} for additional details). We set $\delta^{\ast} = 0.8$, $b = 0.5$, and $a = 0.2$ for all analyses, values chosen to be consistent with empirical estimates of inbreeding depression (e.g., fig.~2 in \citealt{HusbandSchemske1996}). Second, we explore the effects of individual inbreeding depression terms in isolation. That is, we assume that only one of the terms $\delta$ and $\delta_i$ (where $i \in \{J,A,\gamma\}$) can be non-zero at a time.

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Demographic and fitness data for {\itshape M. guttatus}} \label{subsec:MguttMethods}

As an illustrative example of how our model can be used to explore whether demographic rates observed in natural populations appear likely to support balanced SA polymorphisms, we parameterized our model using empirically estimated demographic rates and fitness data for natural populations of the hermaphroditic flowering plant {\itshape Mimulus guttatus} (Scrophulariaceae; now known as {\itshape Erythranthe guttata}). {\itshape M.~guttatus} is an herbaceous, self-compatible wildflower native to western North America that exhibits remarkable among-population variation in numerous life-history and reproductive traits including selfing rates, inbreeding depression, floral morphology, and annual-to-perennial life-history \citep[e.g.,][]{RitlandGanders1987, Ritland1990, Willis1993, Willis1999a, Willis1999b, WuWillis2008}. Moreover, detailed demographic studies have been conducted on multiple populations of {\itshape M.~guttatus}, with demographic matrix data available on the public demographic database COMPADRE \citep{CompadreDB2020}. Below, we briefly outline how we parameterized our model using the available data; full details are provided in \hl{Appendix B.x}.

We used extensive demographic data reported in a large-scale study of local adaptation using experimental populations of {\itshape M.~guttatus} in Stanislaus National Forest (California, USA) in 2012 and 2013 \citep{PetersonEtAl2016}. We leverage their common-garden experimental design to focus on a comparison of demographically viable polymorphic parameter space for two experimental populations with contrasting demographic rates. The first was a locally adapted 'Eagle Meadows' population (data from 2012), while the second group was an experimental population composed of multiple non-locally adapted 'low-elevation perennials' (data from 2013). The vital rate estimates for the Eagle Meadows population are as follows: seed bank survival ($D = 0.534$), seed germination rate ($G = 0.469$), flower production ($F = 0.64$), ovules per flower ($O = 614$), seedling recruits proportional to clonal rosette recruits ($A = 6.7 \times 10^{-4}$), overwinter survival ($S = 0.179$), rosette production ($R = 8.71$). The corresponding estimates for the low-elevation perrenials are: $D = 0.534$, $G = 0.652$, $F = 0.4.09$, $O = 494$, $A = 6.7e-4$), $S = 0$, and $R = 0$ (see corrected Tables.~1 and S2 in \citealt{PetersonEtAl2017}). The same estimates for $D$ and $A$ were used for all populations. The resulting transition matrices for this population involved three life-history stages ($\omega = 3$; seed, seedling, and rosette), and individual elements of the transition matrix ($\tilde{\mbf{A}}$) were calculated as products of the above rates (see Matrix $1$ in \citealt{PetersonEtAl2016}).

Estimates of selfing rates and inbreeding depression were not available for the same experimental populations, but are available for a variety of other western USA {\itshape M.~guttatus} populations. Selfing rate estimates vary in magnitude from near complete outcrossing to predominant selfing ($C \approx 0$ to $0.75$; \citealt{RitlandGanders1987, Ritland1990, Willis1999b}). Estimates of inbreeding depression at several of the life-history stages/fitness components that were included in the data of \citet{PetersonEtAl2016} are available for two intensively studied populations in the Cascade Mountains of Oregon (Iron Mountain and Cone Peak) \citet{Willis1993, Willis1999a, Willis1999b}. Using the data provided in \citet{Willis1993}, we estimated the proportional decrease due to inbreeding depression in seed germination rate ($\delta_{G} = 0.085$), flower number ($\delta_{F} = 0.2$), and overwinter survival ($\delta_{S} = 0.38$). The largest field-estimated selfing rate for this same Iron Mountain population was $C = 0.29$ \citep{Willis1993}.

Using these combined demographic rates, selfing rate, and inbreeding depression estimates, we constructed a corresponding stage $\times$ genotype mendelian matrix model with a single SA locus affecting female and male fertility (as described above). Using this empirically parameterized model, we are able to make predictions about the scope for demographically viable SA polymorphism at a single locus in hypothetical populations with the same demographic rates as observed in \citet{PetersonEtAl2016}, for a range of selfing and inbreeding depression rates observed in other natural populations. We stress, however, that these illustrative predictions are not explicit predictions of the likelihood of SA polymorphism in any specific population.

Interestingly, a polymorphic chromosomal inversion (inv6) with apparently SA fitness effects has been identified in the Iron Mountain population of {\itshape M.~guttatus} \citep{LeeKelly2015}. inv6 segregates at moderate frequency (about $8\%$), and carriers suffer $\approx 30 \%$ loss in pollen viability, but also increased flower production which varies among years. The genetic basis of the fitness effects caused by inv6 are not yet known, but the net result is a "supergene" with strong fitness effects that segregates as a single diallelic locus. We estimated dominance and selection coefficients for the effect of inv6 on pollen viability and flower production from the data reported in \citet{LeeKelly2015}. Across four mapping populations, the effect of inv6 on pollen viability was partially recessive on average ($h_m \approx 0.35$). Given this average dominance, the selection coefficient in the field is approximately $s_m = 0.88$. It was not possible to estimate the average dominance of the effect of inv6 on flower production (inv6 homozygotes were too rare in the field), so we conservatively assume that it was additive ($h_f = 0.5$). Based on this assumption, the selection coefficient inv6 on flower production in the field was $s_f = 0.384$ in 2012, and $s_f = 0.232$ in 2013. As a final proof-of-concept test of our empirically parameterized model \citep[e.g.,][]{Servedio2014}, we asked whether, given the above biologically grounded fitness effect estimates, inv6 appears to fall in demographically viable SA polymorphic parameter space, as might be expected given its observed frequency in the Iron Mountain population.

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}\label{sec:Results}
%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polymorphism and Extinction}\label{subsec:PolyExt}

%%%%%%%%%%%%%%%%%%%%%
% Figure 1
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{../output/figs/extinctionThresholdsFig.pdf}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the mendelian stage-structured matrix model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). "Demographically viable polymorphic parameter space" corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity, where $h = 1/2$, and dominance reversal, where $h = 1/4$); extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel).}}
 \label{fig:extThresholds}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

We begin with a simple illustration of demographically viable polymorphic parameter space in the absence of inbreeding depression (i.e., $\delta_i = 0$). Invasion conditions for SA alleles in the evolutionary demographic model closely match the predictions from population genetic models (see \hl{Online Appendix X} \citealt{Kidwell1977, JordanConnallon2014, Olito2017}). In particular, the demographic model recovers the classic "funnel-shaped" region of polymorphic $s_f \times s_m$ parameter space. The effects of the population selfing rate ($C$) and dominance ($h$) on SA polymorphism are also similar: self-fertilization ({\itshape i}) skews the SA polymorphic space towards regions with weaker selection through female-function and reduces the total area of polymorphic parameter space; and ({\itshape ii}) dominance reversals (where deleterious SA fitness effects are partially recessive in each sex; $h = 1/4$) are much more permissive of SA polymorphism \citep{JordanConnallon2014, Olito2017}.

However, a key prediction from the evolutionary demographic model is that large fractions of SA polymorphic parameter space can be demographically inviable (Fig.~\ref{fig:extThresholds}). The location of the extinction threshold, where the population intrinsic growth rate $\lambda = 1$, is primarily determined by the fertility parameter ($f$) but is also influenced by the population selfing rate ($C$) and dominance of the SA alleles ($h$). When $f$ is large, the population can sustain higher mortality caused by stronger selection through both male and female fitness components before extinction occurs (the extinction threshold shifts to higher values of $s_f$ and $s_m$).

The demographic effects of SA polymorphism are driven by the fitness effects of segregating male-beneficial/female-deleterious alleles because they directly influence individuals' fecundity. For the fertility parameter values we explored, all extinctions occured in regions where either the female-deleterious allele segregates at intermediate frequency (i.e., where SA polymorphism is maintained), or goes to fixation (fig.~\ref{fig:extThresholds}). The SA polymorphisms that remained viable at lower fertilitiy values corresponded to regions where the female-deleterious allele is predicted to segregate at low frequencies, or in regions with weak selection through both sex functions.
